Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-25 drafts/hybrid_zone.tex
some more minor changes

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model.}
\plr{(I presume it's been described, but I don't think we found a reference that actually did this, right?  So maybe say it's ``well-known in the field'' or something?  Also, no-one seems to have looked at lineages, and as our final point says, we actually get some more information out of this perspective.)} \alisa{So, when I sati's been previously described, I'm thinking about the traditional cline theory for a single locus (e.g. Slatkin, May), which describes the time-forward diffusion.}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p +  2 p (1-p)  \times s (\frac{p}{2} - 1), %YB Rearranged this equation to allow for cleaner description
%    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
\yb{where $ \frac{d}{dt} p$ is the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$), and  $\Delta p$ is the Laplacian\alisa{, i.e. the second derivative of allele frequency over space ($x$), at time $t$.}
The first term,  $\sigma^2/2 \Delta p$, describes the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$). 
The second term,  $2 p (1-p)  s (p/2-1)$, describes the impact of selection on local allele frequency as the product in local genotypic variance, $2 p (1-p)$, and selection on allele $A$, $s (p/2-1)$. The term $s (p/2-1)$ clearly describes  underdominance --- $A$ is favored when common ($p>.5$) and disfavored when rare ($p<.5$).}      
%YB: Pardon the extreme changes here, I tried to make things more obvious to a biologist
%and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
%due to local proportions of heterozygotes,
%which is $s(2p-1)$ (up to an error of size $s^2$).
%This can be derived as in \citet{Nagylaki1975};
%for discussion of diploidy see \citet{christiansen1995genotypic}.

@@

Power will be optimal at intermediate values of $s$.
If selection is too strong, it may be difficult to observe the signal
due to a lack of introgressed selected sites,
while if selection is weak, the selected lineage does not move very fast,
and so the strength of signal from elevated $l_B(m)$ is diminished. 
Indeed, if $s$ is less than $x/(\sigma T)$, then only one block of ancestry $A$ is expected to be seen about a selected $A$ allele,
and $l_A(x,m)$ is expected to be $2/T$.
For similar reasons, power and resolution are best at intermediate $T$.

>>>>

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{Equations describing the forward-time establishment of the cline have been previously described \cite[e.g.][]{Slatkin1973,May1975,Nagylaki1975}}
\plr{(I presume it's been described, but I don't think we found a reference that actually did this, right?  So maybe say it's ``well-known in the field'' or something?  Also, no-one seems to have looked at lineages, and as our final point says, we actually get some more information out of this perspective.)} \alisa{So, when I sati's been previously described, I'm thinking about the traditional cline theory for a single locus (e.g. Slatkin, May), which describes the time-forward diffusion.}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p +  2 p (1-p)  \times s (\frac{p}{2} - 1), %YB Rearranged this equation to allow for cleaner description
%    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
\yb{where $ \frac{d}{dt} p$ is the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$), and  $\Delta p$ is the Laplacian\alisa{, i.e. the second derivative of allele frequency over space ($x$), at time $t$.}
The first term,  $\sigma^2/2 \Delta p$, describes the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$). 
The second term,  $2 p (1-p)  s (p/2-1)$, describes the impact of selection on local allele frequency as the product in local genotypic variance, $2 p (1-p)$, and selection on allele $A$, $s (p/2-1)$. The term $s (p/2-1)$ clearly describes  underdominance --- $A$ is favored when common ($p>.5$) and disfavored when rare ($p<.5$).}      
%YB: Pardon the extreme changes here, I tried to make things more obvious to a biologist
%and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
%due to local proportions of heterozygotes,
%which is $s(2p-1)$ (up to an error of size $s^2$).
%This can be derived as in \citet{Nagylaki1975};
%for discussion of diploidy see \citet{christiansen1995genotypic}.

@@

Power will be optimal at intermediate values of $s$ \alisa{relative to $\sigma$}.
If selection is too strong, it may be difficult to observe this signal
due to a lack of introgressed selected sites,
while if selection is weak, the selected lineage does not move very fast \alisa{back in time},
and so the strength of signal from elevated $l_B(m)$ is diminished \alisa{(Fig.~\ref{sfig:blockLengthPlot_t1000_s0.001})}. 
Indeed, if $s$ is less than $x/(\sigma T)$, then only one block of ancestry $A$ is expected to be seen about a selected $A$ allele,
and $l_A(x,m)$ is expected to be $2/T$.
For similar reasons, power and resolution are best at intermediate $T$.
